---
title: "income inequlaity"
author: Naveed Ahmad
execute: 
  echo: false

---

```{r}
#| message: false
#| warning: false
library(tidycensus)
library(tidyverse)
library(tidymodels)
library(broom)
library(ggrepel)
library(plotly)
library(marginaleffects)
```

```{r}
#| message: false
#| warning: false
multi_vars <- c(
  gini = "B19083_001",
  median_income = "B19013_001",
  poverty = "B17001_002")

  socio_data <- get_acs(
  geography = "state",
  variables = multi_vars,
  year = 2020,
  survey = "acs5",
  output = "wide")

  socio_clean <- socio_data |> 
  transmute(
    state = NAME,
    gini = giniE,
    median_income = median_incomeE,
    poverty = povertyE)
```

## Explanatory Plot
```{r}
#| message: false
#| warning: false
# Build interactive plot
plot_ly(socio_clean,
  x = ~median_income,
  y = ~gini,
  type = "scatter",
  mode = "markers",
  text = ~paste(
    "<b>State:</b>", state,
    "<br><b>Gini Index:</b>", round(gini, 3),
    "<br><b>Median Income:</b> $", formatC(median_income, format = "d", big.mark = ""),
    "<br><b>Poverty:</b>", formatC(poverty, format = "d", big.mark = ",")),
  hoverinfo = "text",
  marker = list(
    size = ~sqrt(poverty) / 40,  # Adjusted for visual scale
    color = ~gini,
    colorscale = "Blues",
    showscale = TRUE,
    line = list(width = 1, color = "#FFFFFF"))) %>%
  layout(
    title = "Income Inequality vs Median Income & Poverty (Interactive)",
    xaxis = list(title = "Median Household Income (USD)"),
    yaxis = list(title = "Gini Index"),
    hoverlabel = list(bgcolor = "white", font = list(size = 12))) 
      
```

## Description
<div style="text-align: justify;">The bubble chart visualizes the relationship between median household income and the Gini Index —an established measure of income inequality— in U.S., based on ACS 2016–2020 data. Each bubble represents a state, with its horizontal position indicating the median household income (in USD) and its vertical position showing the Gini Index value. Bubble size  reflects median income, gini index, and povertly level in each state, while color intensity (shades of blue) corresponds to Gini Index values, with darker hues representing lesse income inequality.

This chart provides a clear and interactive way to understand that income inequlaity was prevelent in U.S in 2020 as per ACS 2020 data.
</div> 

```{r}
pov <- socio_clean |>
  select(poverty) |>
  pull() 
log_pov <- log10(pov)

# Plot histogram of transformed data
hist(log_pov,
     main = "Histogram of Log10(Poverty Counts)",
     xlab = "Log10(Number of People in Poverty)",
     col = "lightgreen",
     border = "black")
```

## Model
The liner regression model is:
$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_n X_n + \epsilon
$$

with $\epsilon \sim \mathcal{N}(0, \sigma^2)$.


The estimated linear regression model is:
$$
\hat{gini} = 0.65 - 0.03 \cdot \log(\text{median\_income}) + 0.01 \cdot \log(\text{poverty})
$$


```{r}
income_ineq_model <- linear_reg(engine = "lm") |> 
   fit(gini ~ log(median_income) + log(poverty), data = socio_clean) |>
   tidy(conf.int = TRUE)  |>
  select(term, estimate, conf.low, conf.high)
print(income_ineq_model)
```

## Interpretation: 
The fitted linear regression model examines the relationship between income inequality (as measured by the Gini index) and two predictors: median income and poverty rate. The intercept of the model is 0.651. The coefficient for median income is -0.0284, indicating that, holding poverty constant, a 1% increase in median income is associated with an approximate 0.028 decrease in the Gini index. This effect is statistically significant (p = 0.021), and the 95% confidence interval (–0.0523 to –0.0045) does not include zero, reinforcing the evidence of a negative association. Conversely, the coefficient for poverty is 0.00984, suggesting that, controlling for income, a 1% increase in poverty corresponds to a 0.0098 increase in the Gini index. This relationship is also statistically significant (p = 0.0002), with a confidence interval of 0.00492 to 0.0148. Overall, the model indicates that higher median income is linked to lower inequality, while higher poverty is associated with greater inequality.

```{r}
p <- ggplot(income_ineq_model, aes(x = term, y = estimate)) +
  geom_point(color = "steelblue", size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1, color = "darkgray") +
  labs(title = "Regression Coefficients with 95% Confidence Intervals",
       x = "Term", y = "Estimate") +
  theme_minimal()

# Make it interactive
ggplotly(p, tooltip = c("x", "y")) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

